---
title: "Reproducible Research Project - 2"
author: "Dr.Senthil"
date: "15/07/2019"
output: 
  md_document:
    variant: gfm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Synopsis

The goal of the Project is to explore the NOAA Storm Database and explore the effects of severe weather events on both population and economy.The database covers the time period between 1950 and November 2011.

The following analysis investigates which types of severe weather events are most harmful on:

Health (injuries and fatalities)
Property and crops (economic consequences)


## Data Processing

The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. You can download the file from
 https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2 
 
``` {r}
library("data.table")
library("ggplot2")

download_URL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"

download.file(download_URL, destfile = paste0(getwd(), '/repdata%2Fdata%2FStormData.csv.bz2'))

stormDF <- read.csv("repdata%2Fdata%2FStormData.csv.bz2")
stormDT <- as.data.table(stormDF)

```

## Column Names for examining so that we shall remove unwanted fields

```{r}
colnames(stormDT)

columns2Remove <- colnames(stormDT[, !c("EVTYPE"
  , "FATALITIES"
  , "INJURIES"
  , "PROPDMG"
  , "PROPDMGEXP"
  , "CROPDMG"
  , "CROPDMGEXP")])

# Removing columns
stormDT[, c(columns2Remove) := NULL]

# Recreating StormDT Matrix with only the required columns   
stormDT <- stormDT[(EVTYPE != "?" & 
             (INJURIES > 0 | FATALITIES > 0 | PROPDMG > 0 | CROPDMG > 0)), c("EVTYPE"
                                                                            , "FATALITIES"
                                                                            , "INJURIES"
                                                                            , "PROPDMG"
                                                                            , "PROPDMGEXP"
                                                                            , "CROPDMG"
                                                                            , "CROPDMGEXP") ]


# Making the Columns Cleaner

cols <- c("PROPDMGEXP", "CROPDMGEXP")
stormDT[,  (cols) := c(lapply(.SD, toupper)), .SDcols = cols]

# Make property damage alphanumeric values to numeric values.
propdmgkey <-  c("\"\"" = 10^0,
                 "-" = 10^0, 
                 "+" = 10^0,
                 "0" = 10^0,
                 "1" = 10^1,
                 "2" = 10^2,
                 "3" = 10^3,
                 "4" = 10^4,
                 "5" = 10^5,
                 "6" = 10^6,
                 "7" = 10^7,
                 "8" = 10^8,
                 "9" = 10^9,
                 "H" = 10^2,
                 "K" = 10^3,
                 "M" = 10^6,
                 "B" = 10^9)

# Making crop damage alphanumeric values to numeric values
cropdmgkey <-  c("\"\"" = 10^0,
                "?" = 10^0, 
                "0" = 10^0,
                "K" = 10^3,
                "M" = 10^6,
                "B" = 10^9)

stormDT[, PROPDMGEXP := propdmgkey[as.character(stormDT[,PROPDMGEXP])]]
stormDT[is.na(PROPDMGEXP), PROPDMGEXP := 10^0 ]

stormDT[, CROPDMGEXP := cropdmgkey[as.character(stormDT[,CROPDMGEXP])] ]
stormDT[is.na(CROPDMGEXP), CROPDMGEXP := 10^0 ]

# Calculating Cost for Property and Crop Damages 
stormDT <- stormDT[, .(EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, propcost = PROPDMG * PROPDMGEXP, CROPDMG, CROPDMGEXP, cropcost = CROPDMG * CROPDMGEXP)]

```

## Total Property and Crop Damages cost

```{r}
totalcostDT <- stormDT[, .(propCost = sum(propcost), cropcost = sum(cropcost), Total_Cost = sum(propcost) + sum(cropcost)), by = .(EVTYPE)]

totalcostDT <- totalcostDT[order(-Total_Cost), ]

totalcostDT <- totalcostDT[1:10, ]

```


## Total Fatalities and Injuries

```{r}
totalInjuriesDT <- stormDT[, .(FATALITIES = sum(FATALITIES), INJURIES = sum(INJURIES), totals = sum(FATALITIES) + sum(INJURIES)), by = .(EVTYPE)]

totalInjuriesDT <- totalInjuriesDT[order(-FATALITIES), ]

totalInjuriesDT <- totalInjuriesDT[1:10, ]

```

## Results

## Chart - 1, Top Fatalities

```{r}
Fatalities_DT <- melt(totalInjuriesDT, id.vars="EVTYPE", variable.name = "Types")

ggplot(Fatalities_DT, aes(x=reorder(EVTYPE, -value), y=value))+labs(title = "Top 10 Fatalities", x="Fatalities",y="Frequency")+ geom_bar(stat="identity", aes(fill=Types), position="dodge")+theme(axis.text.x = element_text(angle=45, hjust=1))
```


## Chart - 2, Economic Cost 

```{r}
economic_cost <- melt(totalcostDT, id.vars="EVTYPE", variable.name = "Cost_Type")

ggplot(economic_cost, aes(x=reorder(EVTYPE, -value), y=value))+labs(title = "Top 10 Events for Economic Cost", x="Event Type",y="Cost($)")+ geom_bar(stat="identity", aes(fill=Cost_Type), position="dodge")+theme(axis.text.x = element_text(angle=45, hjust=1))
```
